\documentclass[a4paper]{article}
\usepackage[simple]{dmvn}

\begin{document}
\section{Science}

Science is important to most people living in the modem world for a number of
reasons. In particular, science is important to world peace and understanding,
to the understanding of technology, and to our understanding of the world.
Science is important to world peace in many ways. On one hand scientists have
helped to develop many of the modern tools of war. On the other hand, they have
also helped to keep the peace through, research which has improved life for
people. Scientists have helped us understand the problem of supplying the world
with enough energy; they have begun to develop a number of solutions to the
energy problem~--- for example, using energy from the summd from the atom.

Scientists have also analysed the world's resources. We| can begin to learn to
share the resources with the knowledge provided to us by science. Science
studies the Universe and how to use its possibilities for the benefit of men.
Science is also important to everyone who is affected by modern technology. Many
of the things that make our lives easier and better are the results of advances
in technology and, if the present patterns continue, technology will affect us
even more in the future than it does | now. In some cases, such as technology
for taking salt out of ocean water of technology may be essential for our lives
on Earth.

The study of science also provides people with an understanding of
natural worlds. Scientists are learning to predict earthquakes, are continuing
to study many other natural events such as storms. Scientists are also studying
various aspects of human biology and the origin and developments of the human
race. The study of the natural world may help improve life for many people all
over the world. A basic knowledge of science is essential for everyone. It helps
people to find their way in the changing world.
\end{document}
