\documentclass[a4paper]{article}
\usepackage[simple]{dmvn}

\begin{document}
\cent{{\Large <<Economy and Finance>>\\The Financial, Economic and Trade Division of the USA\par}}

The Financial, Economic and Trade Division monitors developments and changes in U.S. economic, financial
and foreign trade policy. It strives to provide sound basic conditions for Swiss-U.S. trade and
investment relations. The Embassy's economic team keeps track of relevant regulations and legislation
and supports and protects Swiss economic interests in general and in particular. The Division also
covers issues such as agricultural, environmental and competition policies as well as development
cooperation.

The role of the Embassy is to contribute toward further strengthening the excellent bilateral
relationship between Switzerland and the United States. By establishing a Joint Economic
Commission (JEC) in January 2000, both governments institutionalized their already close and
long-standing dialogue. The Commission facilitates periodic high-level government and private sector
exchanges of views on a broad range of multilateral and bilateral economic, commercial and financial
issues of common interest (see 2001 action plan).

The U.S. positions concerning the World Trade Organization (WTO), the Organization for Economic
Cooperation and Development (OECD), the World Bank Group, the International Monetary Fund (IMF)
and the Interamerican Development Bank (IDB) as well as U.S. bilateral and economic relations
are of paramount importance to all countries, especially to highly developed ones such as Switzerland.

Furthermore, the Division closely follows U.S. financial markets and their regulations, macroeconomics,
as well as monetary and fiscal policies.

The Division also supports the promotion of Switzerland as a location for foreign investments.

Based on a new law on trade promotion aimed at strengthening the support network and services for
Swiss exporters, a Swiss Business Hub (SBH) was established at the Swiss Consulate General in Chicago
in October 2001.

Within the framework of the OSEC Business Network Switzerland, the Swiss Business Hub, together with
the Trade Commissioners located at the Swiss Consulates in Atlanta, Chicago, Houston, Los Angeles,
New York and San Francisco--and in cooperation with the Financial, Economic and Trade Division--assists
Swiss companies and Swiss industry in establishing contacts with U.S. partners. Promoting the export
of Swiss products and services demands a multifaceted effort by all the various players.
\end{document}
